---
title: "DASM Practical 3"
author: "Chris Davis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Getting started

Make sure to load the `ggplot2` library so we can do plotting.
```{r}
library(ggplot2)
```

# Hypothesis testing
(step them through this example)

parrot example from lecture (slide 20/50)

first calculate t-statistic
calculate rejection region for alpha 0.5, have to go to limit 0.95

calculate p-value

show this the same with lower tail=FALSE

show how to get from the table and also from R.

(students on their own)
should plot t distribution (19 deg of freedom, alpha 0.05)
add vertical lines to indicate where their own t-values are and where the rejection regions start

show graphical solution, but not code

students do a 2-tailed test on their own (slide 20/50).  
question is if the lifetime is different 

critical region will be different and may also have to reject it.

# First section: t-test

## Summary of the lecture

t test can be used: 
 
* if sample size >~ 30 (distribution does not matter)
* if sample size <~ 30, only if distribution is normal 


## 4 types of t-test in the lecture: (lower tail, upper tail, two tailed)

```{r}
non_smoke = c(18,22,21,17,20,17,23,20,22,21)
smoke = c(16,20,14,21,20,18,13,15,17,21)
```


### one sample t-test
#### Smokers

looking at smokers only, know that in the general population people on average score 20 on the test

```{r}
qqnorm(smoke)
qqline(smoke)
```

Our null hypothesis is that smokers also score 20.  Our alternative hypothesis is that they score less. 

Now we need to do a t-test to see if we can reject the null hypothesis that the smokers also score 20 (or even more)

The syntax of the t-test is `t.test`, type in the RStudio console `?t.test` to see all options

### two sample t-test

```{r}
qqnorm(non_smoke)
qqline(non_smoke)
```

mean is different

Calculate $\bar{x}$

```{r}
x_bar2 = mean(smoke)
x_bar1 = mean(non_smoke)
print(x_bar2)
print(x_bar1)
```

SD is different

```{r}
s2 = sd(smoke)
s1 = sd(non_smoke)
print(s2)
print(s1)
```


The syntax for the t-test command in R is:
```{r, eval=FALSE}
t.test(x, 
       alternative = c("two.sided", "less", "greater"),
       mu = 0, paired = FALSE, var.equal = FALSE,
       conf.level = 0.95)
```

Whenever you see something like `alternative = c("two.sided", "less", "greater")` that means that you should choose one of the values like `alternative = "two.sided"`


For our t-test:

* `x` is smokers
* our `alternative` is `"less"`
* the null hypothesis is that `mu=20`.
* We leave out these other variables like `paired` and `var.equal`


```{r}
t.test(smoke, alternative = "less", mu=20, conf.level = 0.95)
```

Now on your own, try to do a two-sided test and see what happens if you actually make it so the alternative is not less but greater.  You should see the following results below if you did things correctly.

The p-value is 0.9873 since you confused the sidedness of the t-test.  The sample average is actually less than the hypothesized mean.  This calculates the p-value of the upper side.

plot the t-distribution with alpha = 0.05 and degrees of freedom is 9 (since n-1).  For the upper tail t-test, your t value is - point something since it's on the left side of the distribution.  The upper tail t-test calculates the probability of greater than this t-value

Note the negative t value.  The alternative hypothesis is that the true mean is greater than 20, therefore R calculates the p-value as the probability of t > -2.6769, and this probability is large!

```{r, echo=FALSE}
t_test_result = t.test(smoke, alternative = "greater", mu=20, conf.level = 0.95)
print(t_test_result)

```

show the t-distribution, then the actual t-value which was calculated.  (on my example shade the right hand side, students have code without shading)


if you have a sample mean that is lower than your hypothesized mean, you should never do the upper tail test, only do the lower-tail or two-sided test.  The reverse is true if you have a sample mean that is larger than your hypothesis.

Therefore, the two-sided test is a good choice since you can't go wrong.

```{r}
val$statistic
```

will cover this in a later lecture, but now we give you the commands that you can use.

Instead of comparing this set of smokers to the general population, we want to do a more detailed test.  

(slide 48/53)

#### two sample t-test - non-equal variance 

still `alternative` is greater or lesser

alternative greater/lesser corresponds to the mean

null hypothesis goes for the difference in the means, so mu=0 means no difference in the means

this test then calculates the difference in the means = mean of first sample minus the mean of the second sample

pay attention to order of x and y, and what `alternative` you choose

#### two sample t-test - equal or unequal variances
now `var.equal` is important 

`var.equal` specifies if both samples have the same variance.  If you think they come from distributions with the same variance.  If you don't know, then it's better to fill it in as `FALSE`.  If they are the same, then your test is a bit less efficient, but if the variances are different and you fill in `TRUE` then you will get not good results.

(students try to put this into a t-test - see what happens with equal and non-equal variances - see what happens to the p-value so see that they're slightly different)

Look at the mean
```{r}
mean(smoke)
mean(non_smoke)
```
based on the means, can choose which alternatives to take

```{r}
t.test(non_smoke, smoke, alternative = "greater", var.equal = FALSE, conf.level = 0.95)
t.test(non_smoke, smoke, alternative = "greater", var.equal = TRUE, conf.level = 0.95)
```

#### two sample t-test - paired

here `paired` is important

the previous test with the smokers maybe wasn't good - our test could be messed up by chance.

What people often do is have the same people in the test - first take test before they smoke, and then take the test again after they smoked.  If we assume that these values are measured on the same individual, then we have to choose these paried alternatives (this will be explained more in Lecture 4)

```{r}
pre_test = c(77, 56, 64, 60, 57, 53, 72, 62, 65, 66)
post_test = c(88, 74, 83, 68, 58, 50, 67, 64 ,74 ,60)

t.test(pre_test, post_test, alternative = "less", var.equal = FALSE, paired = TRUE, conf.level = 0.95)
t.test(pre_test, post_test, alternative = "less", var.equal = TRUE, conf.level = 0.95)
```


## power of the test 
for the previous example, let them play with the power of the test 
(see power.R for example)

`power.t.test`

```{r, eval=FALSE}
power.t.test(n = NULL, delta = NULL, sd = 1, sig.level = 0.05,
             power = NULL,
             type = c("two.sample", "one.sample", "paired"),
             alternative = c("two.sided", "one.sided"),
             strict = FALSE, tol = .Machine$double.eps^0.25)

```



* `delta` parameter is the true difference between the null hypothesis and the alternative hypothesis.
* `n` is the sample size
* `sd` is the estimated standard deviation of the population
* `sig.level` is the significance level that you choose, e.g. 0.05, (Type I error probability)
* `power` is the power of test (1 - Type II error probability)
* `alternative` - `two.sided` or `one.sided`
* `type` we will only do this for the `one.sample`

We can explore the different alternative hypotheses

always fill in the sig.level and sd.  For n, delta and power, you have to fill in two of them and the test will calculate the remaining one.  If say you know sample size, and the sig.level, it will give you the `power` of your test.  If know the power you want, then this will give you the sample size to achieve this power.

Back to the parrots
(copy/paste from slides)
slide 30/50 only copy upper half (only question)


```{r}
x0 = 28
xa = seq(28.5, 33, by = 0.5)
n = 20


sigma = 5
alpha = 0.05

diff = xa - x0

delta = abs(x0-xa)/sigma*sqrt(n)

res = power.t.test(delta = diff, sd = sigma, sig.level = alpha, n = n, 
                   type = "one.sample", alt = "one.sided")

beta1 = 1- res$power

frame = data.frame(xa, power = res$power)

plot1 = ggplot(frame, aes(xa, power)) + geom_point(aes(size = 2)) + ggtitle("n = 20") +
        theme_classic() + theme(axis.text = element_text(size = 18, color = "black")) + 
        theme(axis.title = element_text(size = 20, color = "black")) + 
        theme(legend.position="none")

```

(slide 31) 
(students explore different hypotheses from 29 to 33, then plot the power of the test as a function as this alternative)
(show the result but not the code)

mu_a is 30, then show the power as a function of sample size (32/50)

should specify the power of 0.9. and the mu_a is 30 and then find the alternative to calculate the power.

# Practical part – bootstrap 
(this is optional if we have time)
Lecture 3 - pages 28-30

include the formula from slide 30

sample doesn't follow a nice t-distribution, so we have to get these samples empirically.  
For alpha = 0.05 (slide 30)

(slide 28 shows how to do it)

```{r}

#basic way of bootsrapping a 95% confidence interval for a small sample

#take a small sample from a exponential distribution

### TODO here they get actual 15 values, do histogram and see that it doesn't come from a normal distribution
x = rexp(rate = 2, n = 15)
hist(x)
### TODO students could also do a q-q plot
#mean and standard deviation of sample
x_bar = mean(x)
x_std = sd(x)

#### TODO convert to loop?
#resampling 1000 times
D = 1000
y = sample(x,D*length(x),replace = TRUE)
y = matrix(y, ncol = D, byrow = FALSE)

### TODO convert this to a loop
#calculate the mean and standerd deviation for each resampled sample
y_bar = sort(colMeans(y))
# explain apply or do something else
y_std = apply(y,2,sd)

#calculate the t-statistic and sort by size
t = sort((y_bar-x_bar)/(y_std)*sqrt(length(x)))
hist(t)
plot(t)

# calculate 0.025 and 0.975 percentiles
t_star=t[c(25,975)]
conf = x_bar + t_star*x_std/sqrt(length(x))

print(conf)
```


# OLD (keep for reference)
# Explain the t-test command and the options 
# Practical part - t-test:
## One sample test:

Give them a large sample >30 with a certain mean xx 
Make a hypothesis that mu < xx 

then they should do an upper-tailed test
let them use the t.test with upper, two tailed, and lower alternative and observe the p-value  (they might be confused)

let them draw  the t-distribution and add the t-statistic from the sample as vertical line, 

for the upper tail test, the p – values is the area to the left of the line
for the lower tail test, 

lesson: the lower tailed test should never be used if sample mean is smaller than mu   

## One sample test:
For the same test: let them compare the p-value given by the t-test with the p-value from the normal distribution 

take the t-value (this is equivalent to z statistic) and calculate qnorm(t, upper.tail = false), should give a result very close to the p-value by the t-test

## two sample t-test 

2 uncorrelated samples: n ~ 20 from normal distributions with equal variances but different means

Do all three kinds of t-test:

* equal variance
* unequal variance
* paired

what p-values do they observe?
