---
title: "Practical 5"
author: "Chris Davis"
date: "October 3, 2016"
output: 
    html_document:
        toc: true
        number_sections: true

---

# Monte Carlo simulation 

## Short reminder of the lecture

x = f(u,v)

u has uncertainty su 
v has uncertainty sv

Construct normal distribution fn,u around u with standard deviation su 
Construct normal distribution fn,v around v with standard deviation sv

Take a random value u1 from  fn,u and v1 from fn,v  and calculate x1 = f(u1, v1)

Repeat n times

Take average (or median) of all xi (i = 1, …n) 

## Example from the lecture (page 25, for now; might be slightly different)
(Chris, maybe add some explanation)

```{r}
n <- 1000
p <- rnorm(n, mean = 2, sd = 0.5)
q <- rnorm(n, mean = 5, sd = 1)
f <- p*q/(p+q)
f_bar <- mean(f)
sig <- sd(f)
```

## Try on your own:

Give them a different formula  x = f(u,v)

Let them try a monte carlo simulation with different n 

Tasks:
 
* plot the histogram of x
* calculate mean(x) and std(x)

If there is enough time you could explain, how to do it with a loop for different n and re-create a plot like in lecture pgs 27, 28

# Working with real data sets

This is not related to the lecture, so I think I leave this to you. 

I attach two practicals that they had last year, where I taught them the basic of working with data frames, subsets, dates and times etc … 

I think it would be good to keep the structure that we had so far:

* Introduce a concept
* walk them through an example 
* let them try an exercise by themselves




The Guerry data file:

Read it in: 
```{r}
data1 <- read.csv(file ="/Users/uli/_Work/Teaching/Statistics_course/Lecture_6/testdata.csv")
```
Get an idea what is in it:
```{r}
summary(data1)
```

This is a data set from the Guerry R package, so to get an idea about the variables we need to install the package:

```{r}
install.packages("Guerry")
```

and then bring up the help information that relates to this specific data frame which is included in the package:
```{r}
help(Guerry)
```
French cities and statistics about them

A few categories: Region, Main_city
A few numerical variables: e.g., Crime pers … Suicides
A few rank variables: Donations, 


1) Select any numerical variable and plot a histogram
 hint:	  (data1$xxx)

2) Choose variable ”Literacy” and make a box plot per region and city size (MainCity)  

hint: data1~xxx, groups data by the categories contained in the variable xxx


1) Select any numerical variable and plot a histogram
	
hist(data1$Infants)
	

2) Choose numerical variable and make a box plot per region and city size (MainCity)  


boxplot(data1$Literacy~data1$Region)
boxplot(data1$Literacy~data1$MainCity)


3) we see that the box in the small cities looks weird, we want to take a 
closer look 

3a) First we want to know how many variables are in the small, medium and large cities

a <- table(data1$MainCity)

 1:Sm 2:Med  3:Lg 
   10    66    10

3b) Now we want to see the actual data points in the small cities only:
Subset command!


The subset( ) function is the easiest way to select variables and observations. In the following example, we select all rows that have a value of age greater than or equal to 20 or age less then 10. We keep the ID and Weight columns.

 using subset function 
```{r}
newdata <- subset(mydata, age >= 20 | age < 10, select=c(ID, Weight))
```

In the next example, we select all men over the age of 25 and we keep variables weight through income (weight, income and all columns between them).

using subset function (part 2)

```{r}
newdata <- subset(mydata, sex=="m" & age > 25, select=weight:income)
```

3b) Now: apply to data1

```{r}
small_data <- subset(data1, MainCity ==  "1:Sm", select = Literacy)
plot(small_data)
```

4) Make a boxplot of crime vs city size

You see a large outlier in the small cities. This would disturb a test like ANOVA. remove the outlier

data2 <- subset(data1, !(MainCity == “1:Sm” & Crime_prop > 15000))
boxplot(data2$Crime_prop~data2$MainCity)

5) select any row and column you want 

b <- data1[1:3, 5:9]
c <- data1[c(1,5,9), 7]

6) Make a table of city size vs region

table(data1$mainCity,data1$Region)

tab = table(data1$MainCity, data1$Region)
tab

