---
title: "Practical 4"
author: "Chris Davis"
date: "September 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Computer practicum:

# Section 1: The test for equality in variance

## 1.1Reminder from lecture:

A test of s in two different populations with size  $n_1$, $n_2$ involves the following:

1) A confidence level 1-a
2) Two independent, random samples of normal distributions
3) Null hypothesis, H0: $s_1$ ≤ $s_2$, or $s1$ ≥ $s2$, or $s1$ = $s2$
4) Alternative hypothesis, $H_a$: 
s1 > s2     (upper tail alternative)
s1 < s2     (lower tail alternative)
s1 = s2     (two tailed)
5) Test statistic: $F = \frac{{s_{1}}^2}{{s_{2}}^2}$


## 1.2 Syntax in R:
```{r, eval=FALSE}
var.test(x, y, ratio = 1, alternative = c("two.sided", "less", "greater"),
         conf.level = 0.95, ...)
```


## 1.3 Example from the lecture (show and explain the students)

Take 100 samples from two normal distributions with different mean, but the same variance. Will the test recognize that we should not reject the null hypothesis?

x <- rnorm(100, mean=0)  # if you don’t specify the variance it is set to 1
y <- rnorm(100, mean=1)
 
var.test(x, y, ratio = 1, alternative = "two.sided”,
         conf.level = 0.95, ...)      
         
F test to compare two variances

## 1.4 Exercise on you own 
 a) Take 100 random samples from normal distributions with a different variance e.g., mean = 0, sd = 1 and sd = 1.2 
 … run the var.test again
b) try both above examples again with n = 10 


# Section 2: chi-square test for tabular data

## 1.1 Preliminaries constructing tables from categorical data 

reminder about reading in data sets, getting an idea of the variables etc
extract variables with the $ command 
construct a table from one or two variables 
syntax of the table command

## 1.2 Demonstrate on the example with buckled parents and children from lecture 6 (these are from a data set in R, but I need to look up on Monday, which one it was – book is at the office now)

## 1.3 Exercise:
Let them construct a two way table from other categorical variables

## 2 chi-square test for goodness of fit

### 2.1 Reminder from the lecture

Hypothesis: n events that occur with probabilities pi

Observations: counts for each event (n1, n2, …, ni, …,)

Test statistic:  

$$ x^{2} = \sum{}\frac{(observed - expected)^{2}}{expected} = \sum_{i=1}^{k} \frac{(n_i-E_i)^2}{E_i}$$

with $E_i = n \pi$

### 2.2 Syntax of chi-square test

chisq.test(x, p =) ** note to chris, I don’t want to introduce the other options, they can ignore them

p should be a vector of length x. If it is not specified the program automatically assumes equal probabilities. 

### 2.3 Example 
the dice from the lecture 6

### 2.3 Try on your own:
the example of the murder cases from the lecture 

